Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-06-03 SCP2008/paper/metamonads.tex
A few notes referencing the hand-written GE code

\section{Gaussian Elimination}\label{GA-algorithm}

@@

We start by abstracting out domains. We have already seen the simplest
case of domain abstraction in |param_code1|, which took
code-generators such as |plus| and |one| as arguments.  We need far
more than two parameters: our domains should include $0$, $1$, $+$,
$*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. We could group
these parameters in tuples or records.  Instead, we use OCaml
\emph{structures} (i.e., modules) so we can take advantage of
extensibility, type abstraction and constraints, and especially
parameterized structures (\emph{functors}).  We define the type,
the signature |DOMAIN|, which different domains must satisfy:

@@

The signature |CONTAINER2D| below specifies that a container must provide
functions |dim1| and |dim2| to extract the dimensions, functions |getL|
to generate container getters, the cloning
generator |copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details.  Note that
this module type is also abstract over the actual representation 
|rep| of the generator data-structure.

@@

To track the determinant, we should be able to generate code
for: defining variables used for tracking (|decl|),
updating the sign (|upd_sign|) or the absolute
value of the determinant, and converting the tracking state
to the final determinant value of type |outdet|. LU of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| (see definition below) where |outdet| is |unit|
and all the
functions of that module generate no code. For integer matrices, and
in general whenever the matrix elements do not form a field, we
have to track some aspects of the determinant, even if we do not output
it, as the fraction-free update needs those aspects. The determinant tracking
aspect is complex because tracking variables, if any, are to be declared at the
beginning of LU; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
code for the value of the type |Dom.v ref|, among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state. 
Other aspects follow a similar outline.

@@

The second instance does track the determinant. The |decl| method
generates two let-bindings for mutable variables tracking the
magnitude and the sign of the determinant, and places the names of
these variables in the monadic state. The |upd_sign| method, invoked
from row- or column-swap generators, retrieves the name of the
sign-accumulating variable and generates the code to update the
sign. The |fin| method retrieves the names of both accumulators and
generates code to compute the final, signed determinant value.
\begin{code}
module AbstractDet =
  struct
  open C.Dom
  type tdet = v ref
  type 'a lstate = ('a,int ref) abstract * ('a,tdet) abstract
  let decl () = perform
      magn <-- retN (liftRef oneL);    (* track magnitude *)
      sign <-- retN (liftRef Idx.one); (* track the sign: +1, 0, -1 *)
      mo_extend ip (sign,magn);
      unitL
  let upd_sign () = perform
      (sign,_) <-- mo_lookup ip;
      ret (Some (assign sign (Idx.uminus (liftGet sign))))
  let fin = fun () -> perform
      (sign,magn) <-- mo_lookup ip;
      ifM (Logic.equalL (liftGet sign) Idx.zero) (ret zeroL)
      (ifM (Logic.equalL (liftGet sign) Idx.one) (ret (liftGet magn))
          (ret (uminusL (liftGet magn))))
  \dots
end
\end{code}

@@

The type |wmatrix| is for convenience, a bundling of several pieces
of data which are needed by both the |OUTPUT| aspect and the |PIVOT| 
aspect.  It is defined as
\begin{code}
type 'a wmatrix = {matrix: 'a C.vc; numrow: ('a,int) abstract; 
                   numcol: ('a,int) abstract}
\end{code}

@@

\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)
let instantiate gen =
    .<fun a -> .~(runM (gen .<a>.) []) >.;;
let resIV5 = instantiate GenIV5.gen ;;
\end{code}
%
We first instantiate the container |GVC_I| for a vector over an
integer domain.  |GenGE| is then one functor of many algorithm
generators from a linear algebra library; since all our linear algebra codes
are parameterized by the container and the domain included in it,
We pass the container to the main functor |GenLA|.
We open the resulting |G_GVC_I| module
to make all its components easily available. One of those components
is the GE generator, |GE| (which we open as well). Besides |GE|, the
module |G_GVC_I| includes the generator for GE-based solvers.  The main
generator functor is |GenGE|, shown later.  We instantiate the
generator by passing to the functor the ``record'' of various aspects; the
order is irrelevant, but all aspects such as |Det| must be
specified. In the code above, we request GE generation with full
pivot, fraction-free update, operate on non-augmented matrix and
return the |U| factor, determinant and the rank. We run the monad
passing the initial state |[]| and obtain the code, which we can see
by printing |resIV5|. The code can then be ``compiled'' as |!. resIV5|
or with offshoring\cite{offshoring}. The code for |resIV5|
(App.~\ref{app:code-GenIV5}) shows full
pivoting, determinant and rank tracking. The code for all these
aspects is fully inlined; no extra functions are invoked and no tests
other than those needed by the LU algorithm itself are performed. The
resulting function returns a triple |int array * int * int| of the
U-factor, determinant and the rank.

@@

A careful reading of this code will reveal that the core of the 
Gaussian Elimination algorithm is still visible in this code generator:
the |forward_elim| function iterates over the columns and rows of the
matrix, finding a pivot, and zeroing the appropriate entries.

>>>>

\section{Gaussian Elimination}\label{GE-algorithm}

\oleg{I wonder should we always refer to GE rather than LU-elimination
  in our paper? GE is a particulat instance of LU; one can accomplish
  LU in other ways than GE. Also, GE can be used for other means that
  LU, e.g., just to compute the determinant or to solve linear equations.
  So, it seems more properly to say we do GE. If we use GE and LU
  almost synonymously, as it seems to be the case in Mathematica, we
  should note that in a footnote.}

@@

We start by abstracting out domains. \oleg{Add: clearly the basic
  structure of the algorithm, Fig.~\ref{fig:GE-algorithm} remains the
  same for integer, float, etc. matrices.} We have already seen the
simplest case of domain abstraction in |param_code1|, which took
code-generators such as |plus| and |one| as arguments.  We need far
more than two parameters: our domains should include $0$, $1$, $+$,
$*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. We could group
these parameters in tuples or records.  Instead, we use OCaml
\emph{structures} (i.e., modules) so we can take advantage of
extensibility, type abstraction and constraints, and especially
parameterized structures (\emph{functors}).  We define the type, the
signature |DOMAIN|, which different domains must satisfy:

@@

\oleg{Say that CONTAINER2D is an abstraction of a particular
  container, flat vector, used in the algorithm in
  \S\ref{GE-algorithm}.}
The signature |CONTAINER2D| below specifies that a container must provide
functions |dim1| and |dim2| to extract the dimensions, functions |getL|
to generate container getters, the cloning
generator |copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details (see
\S\ref{GE-algorithm} for an example of such an optimiation).  Note that
this module type is also abstract over the actual representation 
|rep| of the generator data-structure.

@@

To track the determinant, we should be able to generate code
for: defining variables used for tracking (|decl|),
updating the sign (|upd_sign|) or the absolute
value of the determinant, and converting the tracking state
to the final determinant value of type |outdet| (see
Fig.~\ref{fig:GE-algorithm} for an example).
LU of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| (see definition below) where |outdet| is |unit|
and all the
functions of that module generate no code. For integer matrices, and
in general whenever the matrix elements do not form a field, we
have to track some aspects of the determinant, even if we do not output
it, as the fraction-free update needs those aspects. The determinant tracking
aspect is complex because tracking variables, if any, are to be declared at the
beginning of the code (see lines 22-23 (XXX) of
Fig.~\ref{fig:GE-algorithm}); the sign
of the determinant has to be updated on each row or column
permutation (lines 32 and 37 of the figure XXX);
the value of the determinant should be updated per each
pivoting (line 56). We use |lstate| to pass the tracking state, e.g., a piece of
code for the value of the type |Dom.v ref|, among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state. 
Other aspects follow a similar outline.

@@

The second instance does track the determinant. The |decl| method
generates two let-bindings for mutable variables tracking the
magnitude and the sign of the determinant, and places the names of
these variables in the monadic state. The |upd_sign| method, invoked
from row- or column-swap generators, retrieves the name of the
sign-accumulating variable and generates the code to update the
sign. The |fin| method retrieves the names of both accumulators and
generates code to compute the final, signed determinant value.
The generated code closely corresponds to the lines dealing with 
|det_magn| and |det_sign| in Fig.~\ref{fig:GE-algorithm}.
\begin{code}
module AbstractDet =
  struct
  open C.Dom
  type tdet = v ref
  type 'a lstate = ('a,int ref) abstract * ('a,tdet) abstract
  let decl () = perform
      magn <-- retN (liftRef oneL);    (* track magnitude *)
      sign <-- retN (liftRef Idx.one); (* track the sign: +1, 0, -1 *)
      mo_extend ip (sign,magn);
      unitL
  let upd_sign () = perform
      (sign,_) <-- mo_lookup ip;
      ret (Some (assign sign (Idx.uminus (liftGet sign))))
  let fin = fun () -> perform
      (sign,magn) <-- mo_lookup ip;
      ifM (Logic.equalL (liftGet sign) Idx.zero) (ret zeroL)
      (ifM (Logic.equalL (liftGet sign) Idx.one) (ret (liftGet magn))
          (ret (uminusL (liftGet magn))))
  \dots
end
\end{code}

@@

The type |wmatrix| is for convenience, a bundling of several pieces
of data which are needed by both the |OUTPUT| aspect and the |PIVOT| 
aspect.  It is defined as
\begin{code}
type 'a wmatrix = \{matrix: 'a C.vc; numrow: ('a,int) abstract; 
                   numcol: ('a,int) abstract\}
\end{code}

@@

\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)
let instantiate gen =
    .<fun a -> .~(runM (gen .<a>.) []) >.;;
let resIV5 = instantiate GenIV5.gen ;;
\end{code}
%
We first instantiate the container |GVC_I| for a vector over an
integer domain.  |GenGE| is then one functor of many algorithm
generators from a linear algebra library; since all our linear algebra codes
are parameterized by the container and the domain included in it,
We pass the container to the main functor |GenLA|.
We open the resulting |G_GVC_I| module
to make all its components easily available. One of those components
is the GE generator, |GE| (which we open as well). Besides |GE|, the
module |G_GVC_I| includes the generator for GE-based solvers.  The main
generator functor is |GenGE|, shown later.  We instantiate the
generator by passing to the functor the ``record'' of various aspects; the
order is irrelevant, but all aspects such as |Det| must be
specified. In the code above, we request GE generation with full
pivot, fraction-free update, operate on non-augmented matrix and
return the |U| factor, determinant and the rank. We run the monad
passing the initial state |[]| and obtain the code, which we can see
by printing |resIV5|. The code can then be ``compiled'' as |!. resIV5|
or with offshoring\cite{offshoring}. The code for |resIV5|
(App.~\ref{app:code-GenIV5}) shows full
pivoting, determinant and rank tracking. The code for all these
aspects is fully inlined; no extra functions are invoked and no tests
other than those needed by the LU algorithm itself are performed. The
resulting function returns a triple |int array * int * int| of the
U-factor, determinant and the rank. It is instructive to compare the
generated code with the corresponding code in
Fig.~\ref{fig:GE-algorithm}, the hand-written implementation of the
textbook pseudo-code; the only difference is the naming of
variables and the inlining of pivoting and swapping functions.

@@

A careful reading of this code will reveal that the core of the 
Gaussian Elimination algorithm (Fig.~\ref{fig:GE-algorithm}) 
is still visible in this code generator:
the |forward_elim| function iterates over the columns and rows of the
matrix, finding a pivot, and zeroing the appropriate entries.
